Binning Strategies and Related Loss for Large Data

### 2015-03-20 Paper.Rnw
Edits to section 4.2

Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus theoretically have no frequency loss. While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.  Color perception of shade is extremely context sensitive (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.

\hh{It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}). Figure~\ref{fig:context} shows an example of the context sensitivity of colors. Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each of the points clouds.
}

@@

Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive.As a result, a binning width which produces fewer than seven frequency categories is preferable, while minimizing numerical loss within that constraint.

 Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information. We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins  using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.

@@

Frequency data consists of counts, which most commonly exhibit skew densities, i.e.\ there are usually a lot of cells with small cell counts and a few cells with extremely large counts. A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. The \textit{log frequency loss}, $L^{\log F}$, is defined as
%
\begin{eqnarray}\label{logfrequency}
L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
\end{eqnarray}
%
Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.

\ktm{Can't figure out how to get rid of the ``washed-out'' look in the continuous color scale figure.  It looks lighter because it is being rendered with tiny gaps between the 1000 narrow tiles which lightens the whole plot because the background is showing through.}

<<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold'>>= 
J= 1
J2 = 200
bound = 1000
id <- seq(0+J/2,bound-J/2,by=J )
fill <- seq(0+J/2,bound-J/2,by=J )
id2 <- seq(0+J2/2,bound-J2/2, by=J2)
fill2 <- seq(J2/2,bound-(J2/2), by=J2)
p1 <- qplot(id, 2, geom="tile", fill=-fill) +
  theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
        legend.position = "none",
        axis.ticks=element_blank(),axis.text.y=element_blank(),
        panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
        panel.background=element_blank()
        ) + 
  ylab("Continuous") + xlab("")
p2 <- qplot(id2, 2, geom="tile", fill=-fill2) + 
  theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
        legend.position = "none",
        axis.ticks=element_blank(),
        axis.text.y=element_blank(),
        panel.background=element_blank()) + 
  ylab("Discrete") + xlab("Count")

@@

\subsection{Loss of Frequency Information} 
Loss of frequency information in a scatterplot happens because of our inability to render elements of a third dimension efficiently. By employing a color scheme for filling tiles according to the observed frequencies corresponds to a binning of the frequencies into a few different levels, with losses stemming from the same source as the previously discussed numerical loss.
We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_k\}$ with $1 \le k \le n_C$. An observed frequency $c_j$ with $1 \le j \le N_x \times N_y$, where $N_x$ and $N_y$ are the number of bin divisions in the X and Y directions respectively,  is assigned to the closest frequency level $c^\ast_k = b(c_j)$.  
The \textit{unscaled} frequency loss is then defined as
\begin{eqnarray}\label{eqn:frequency}
L_\text{Freq} = \sum_{j=1}^{N_x N_y} \left(c_j - b(c_j)\right)^2
\end{eqnarray}

@@

\paragraph{Traditional Scatterplots and Frequency Information:}

>>>>

Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus theoretically have no frequency loss. While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.  Color perception of shade is extremely context sensitive allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.
\hh{It is therefore not realistic, to expect readers to be able to map colors to a continuous color scheme, even though theoretically we certainly can distinguish enough different colors (see e.g.\ the discussion in \url{http://hypertextbook.com/facts/2006/JenniferLeong.shtml}). Figure~\ref{fig:context} shows an example of the context sensitivity of colors. Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each of the points clouds.
}

@@

Research suggests that even under optimal conditions, we can effectively compare only about seven colors \citep{colorperception}, which provides a physical upper limit on the amount of frequency variation we can perceive. As a result, a binning width which produces seven of fewer frequency categories is preferable.%, while minimizing numerical loss within that constraint.

 Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information. We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.  Figure~\ref{fig:FreqLossDemo} provides a visual example of a discrete color palette resulting from frequency binning.

<<FreqLossDemo,echo=F,include=T,eval=T,fig.width=6, fig.height=3, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Continuous and discrete color scales for counts in the range 0 to 1000. Frequency binning done using standard rectangular binning with origin $\\beta_0 = 0$ and binwidth $\\omega_X = 200$. \\ktm{Can't figure out how to get rid of the ``washed-out'' look in the continuous color scale figure.  It looks lighter because it is being rendered with tiny gaps between the 1000 narrow tiles which lightens the whole plot because the background is showing through.}">>= 
J= 1
J2 = 200
bound = 1000
id <- seq(0+J/2,bound-J/2,by=J )
fill <- seq(0+J/2,bound-J/2,by=J )
id2 <- seq(0+J2/2,bound-J2/2, by=J2)
fill2 <- seq(J2/2,bound-(J2/2), by=J2)
p1 <- qplot(id, 2, geom="tile", fill=-fill) +
  theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
        legend.position = "none",
        axis.ticks=element_blank(),axis.text.y=element_blank(),
        panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
        panel.background=element_blank()
        ) + 
  ylab("Continuous") + xlab("")
p2 <- qplot(id2, 2, geom="tile", fill=-fill2) + 
  theme(plot.margin = unit(c(0.1,1,0.1, 1),"cm"), 
        legend.position = "none",
        axis.ticks=element_blank(),
        axis.text.y=element_blank(),
        panel.background=element_blank()) + 
  ylab("Discrete") + xlab("Count")

@@

Frequency data consists of counts, which most commonly exhibit skew densities, i.e.\ there are usually a lot of cells with small cell counts and a few cells with extremely large counts. The use of quantile binning is promising in the case of frequency binning because it seeks to place the same number of values in each frequency bin, thus balance the number of tiles at each shade.

Another alternative is to use a log transformation which produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. The \textit{log frequency loss}, $L^{\log F}$, is defined as
%
\begin{eqnarray}\label{logfrequency}
L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
\end{eqnarray}
%
Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.

@@

\subsection{Loss of Frequency Information (old writeup - to be reworked or removed)} 
Loss of frequency information in a scatterplot happens because of our inability to render elements of a third dimension efficiently. By employing a color scheme for filling tiles according to the observed frequencies corresponds to a binning of the frequencies into a few different levels, with losses stemming from the same source as the previously discussed numerical loss.
We will assume that we have a set of size $n_C$ of  frequency levels $\{c^\ast_k\}$ with $1 \le k \le n_C$. An observed frequency $c_j$ with $1 \le j \le N_x \times N_y$, where $N_x$ and $N_y$ are the number of bin divisions in the X and Y directions respectively,  is assigned to the closest frequency level $c^\ast_k = b(c_j)$.  
The \textit{unscaled} frequency loss is then defined as
\begin{eqnarray}\label{eqn:frequency}
L_\text{Freq} = \sum_{j=1}^{N_x N_y} \left(c_j - b(c_j)\right)^2
\end{eqnarray}

@@

\paragraph{Traditional Scatterplots and Frequency Information: (old writeup - to be reworked or removed)}
